---
title: "RippleR Tutorial (Xenium Object)"
output:
  html_document:
    df_print: paged    
    toc: TRUE
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    
---

# Import libraries
```{r setup, message=FALSE, warning=FALSE}

# Load libraries
library(RippleR); library(Giotto); library(Seurat); library(dplyr); library(data.table)

# parallelization
library(future); 
plan(multisession,
     workers = max(1, parallelly::availableCores() - 2))


# Adjust global/maxSize to prevent future error
options(future.globals.maxSize = 200000*1024^2)

# Set seed for reproducibility
set.seed(42)

# Knit option
knitr::opts_chunk$set(fig.retina = 2)

```

# Import Xenium object
``` {r load object}
# First import Xenium object and extract relevant info

# Load Xenium object with cell annot metadata
xenium.obj <- readRDS("/gpfs/data/heratilab/YeeHoon/XeniumProject/Xenium_Breslow_SLN_StageIvsStageIII_LoadXenium_Integration_complement_Codex/Giotto_Xenium_Analysis/Giotto_Analysis_Version3/Function_Rpackage_Xenium/Xenium.obj.L-annots-coords.rds")

# Check Xenium object
xenium.obj 

# Get metadata
md <- xenium.obj[[]]

# Check metadata
head(md)

```


# Function to generate concentric circles

## Function Definition
``` {r function definition: generate concentric circles, eval = FALSE}
This function generates concentric circles radiating from a tumor center or custom annotated region.
Directional options for expansion include: "left", "right", "top", "down", or "all".
Note: Circle expansion is automatically constrained by the tissue boundary.

generate_concentric_circles_xenium <- function (xenium.obj, sample, L.annot="L3.annot", Tumor = "Tumor.1", conc.radius.start = 200, conc.radius.stepSize =200, conc.direction = c("right", "left", "top", "bottom", "full"), save_output = TRUE, output_file = "concentric.circle.outputs.rds")

# Params:
#' @param xenium.obj A Seurat Xenium object containing cell metadata and
#'   spatial coordinates.
#' @param sample Character string specifying the sample name to analyze.
#'   Must match values stored in `orig.ident`.
#' @param L.annot Character string specifying the metadata column containing
#'   cell annotations.
#' @param Tumor Character string specifying the annotation label identifying
#'   tumor cells.
#' @param conc.radius.start Numeric. Starting radius for the first circle
#'   (default = 200).
#' @param conc.radius.stepSize Numeric. Increment between successive circle
#'   radii (default = 200).
#' @param conc.direction Character string specifying the direction of circle
#'   expansion. One of:
#'   \itemize{
#'     \item{"full"} Full concentric circles.
#'     \item{"right"} Right-facing semicircles.
#'     \item{"left"} Left-facing semicircles.
#'     \item{"top"} Upward-facing semicircles.
#'     \item{"bottom"} Downward-facing semicircles.
#'   }
#' @param save_output Logical. If TRUE, save the results as an RDS file.
#' @param output_file Character string specifying the output RDS filename.


generate_concentric_circles_xenium() function returns a list containing metadata, spatial geometries, and concentric circle information required for downstream analyses.

Output files:
xenium.sub — filtered cell metadata for the selected sample.
concentric_circles_sf — spatial polygons representing concentric circles or semicircles.
center_point_sfc — tumor centroid stored as an sf point object.
radii — vector of generated circle radii.

```

## Deploy function to generate concentric circles
``` {r function deployment: generate concentric circles}

# Deploy the function to generate concentric circles shifting to the right.
generate_concentric_circles_xenium(xenium.obj= xenium.obj, sample ="SLN_stageIII", L.annot="L3.annot", Tumor = "Tumor.1", conc.radius.start = 200, conc.radius.stepSize =200, conc.direction = "right", save_output = TRUE, output_file = "concentric.circle.outputs.rds")


```



# Retrieve saved res.conc RDS file
``` {r retrieve saved res.conc RDS file}

# Retrieve conc circle outputs from saved RDS file
res.conc <- readRDS("concentric.circle.outputs.rds")
# Check list of outputs saved under res.conc
names(res.conc)
# Unpacks outputs into global environment
list2env(res.conc, envir = .GlobalEnv)

```

# Function to plotting concentric circles and cell annotations

## Function Definition
``` {r function definition: plot concentric circles and cell annotations, eval = FALSE}
This function overlays concentric circle boundaries and cell type annotations onto the spatial tissue layout.


plot_concCircle_cellAnnot_xenium <- function(data, concentric_circles_sf, center_point_sfc, L.annot="L3.annot", Tumor = "Tumor.1", tumor.color = "springgreen3", viridis.option =c("magma", "inferno", "plasma", "viridis", "rocket"), save_plot = TRUE, plot_file = "concentricSemiCircle-Tumor-center.png")

  
# Params:
#' @param data Data frame containing spatial coordinates and cell annotation
#'   metadata. Typically \code{res.conc$xenium.sub}.
#' @param concentric_circles_sf sf object containing concentric circle
#'   geometries generated by \code{generate_concentric_circles()}.
#' @param center_point_sfc sf point object representing the tumor centroid.
#' @param L.annot Character string specifying the metadata column used for
#'   cell annotation.
#' @param Tumor Character string specifying the tumor annotation label.
#' @param tumor.color Color used to highlight the tumor annotation.
#' @param viridis.option Viridis palette option. One of
#'   \code{"magma"}, \code{"inferno"}, \code{"plasma"},
#'   \code{"viridis"}, or \code{"rocket"}.
#' @param save_plot Logical; whether to save the generated plot as a PNG file.
#' @param plot_file Output plot as PNG.


plot_concCircle_cellAnnot_xenium() function visualizes spatial cell annotations together with the concentric circles generated by generate_concentric_circles_xenium(). Cells are colored according to the selected annotation column, while the tumor population can be highlighted using a custom color.


Output files:
l3.annot_colors — a named vector mapping each annotation category to a color.
concCircle.plot — a ggplot object containing the spatial visualization.


The generated plot includes:
Spatially resolved cells colored by annotation.
Concentric circles or semicircles surrounding the tumor centroid.
A red marker indicating the tumor centroid.
A custom tumor color for easy visualization of the tumor population.

If save_plot = TRUE, the figure is automatically exported as a high-resolution PNG file. This allows the plot to be reused in manuscripts, presentations, or downstream reports without rerunning the function.


This function is typically used immediately after generate_concentric_circles_xenium() to visually verify tumor centroid placement, circle geometry, and spatial annotation patterns before performing radius-based analyses.


```


## Deploy function to generate concentric circles and cell annotations
``` {r function deployment: plot concentric circles and cell annotations}

res <- plot_concCircle_cellAnnot_xenium(data = res.conc$xenium.sub, concentric_circles_sf = res.conc$concentric_circles_sf, center_point_sfc = res.conc$center_point_sfc, L.annot="L3.annot", Tumor = "Tumor.1", tumor.color = "springgreen3", viridis.option =c("magma", "inferno", "plasma", "viridis", "rocket"), save_plot = TRUE, plot_file = "concentricSemiCircle-Tumor-center.png")


```

## Show concentric circle plot 
``` {r show concentricSemiCircle-Tumor-center plot, fig.width=12, fig.height=10}
# Check list of outputs saved under res
names(res)

# Show plot with concentric circles emanating from Tumor center
res$concCircle.plot

```


# Function to color concentric circles by radius

## Function Definition
``` {r function definition: plot concentric circles with colored radii, eval = FALSE}
This function visualizes concentric circles emanating from the tumor center, color-coded by radial distance.

plot_concCircle_colorRadii_xenium <- function(data, Tumor = "Tumor.1", concentric_circles_sf, center_point_sfc, g.title="Concentric Circles emanating from Tumor Center", scale.viridis.option=c("C", "A", "B", "D", "F"), save_plot = TRUE, plot_file="concentricSemiCircle-Tumor-center_coloredRadii.png")

  
# Params:
#' @param data Data frame containing spatial coordinates. Typically
#'   \code{res.conc$xenium.sub}.
#' @param Tumor Character string specifying the tumor annotation label.
#'   Used for naming the output file.
#' @param concentric_circles_sf sf object containing concentric circle
#'   geometries generated by \code{generate_concentric_circles()}.
#' @param center_point_sfc sf point object representing the tumor centroid.
#' @param g.title Plot title.
#' @param scale.viridis.option Viridis color palette option. One of
#'   \code{"A"}, \code{"B"}, \code{"C"}, \code{"D"}, or \code{"F"}.
#' @param save_plot Logical; whether to save the generated plot as a PNG file.
#' @param plot_file Output filename used when \code{save_plot = TRUE}.


The plot_concCircle_colorRadii_xenium() function generates a spatial visualization of the concentric circles created by generate_concentric_circles_xenium(), with each radius displayed using a distinct Viridis color.

Output file:
concCircle.plot.radius: a single ggplot object, concCircle.plot.radius, which can be displayed directly, further customized using standard ggplot2 syntax, or exported for publication-quality figures. 

When save_plot = TRUE, the plot is automatically saved as a high-resolution PNG file (concentricSemiCircle-Tumor-center_coloredRadii.png by default) for downstream reporting and visualization.

This figure is typically reviewed before performing radius-based analyses such as annotate_radius_bins_xenium(), label_peritumoral_distal_xenium(), and gene_count_radius_xenium(), all of which depend on accurate concentric circle geometry and radius definitions.

```

## Deploy function to color concentric circles by radius
``` {r function deployment: plot concentric circles with colored radii}

res <- plot_concCircle_colorRadii_xenium(data = res.conc$xenium.sub, Tumor = "Tumor.1", concentric_circles_sf = res.conc$concentric_circles_sf, center_point_sfc= res.conc$center_point_sfc, g.title="Concentric Circles emanating from Tumor Center", scale.viridis.option="C", save_plot = TRUE, plot_file="concentricSemiCircle-Tumor-center_coloredRadii.png")


```


# Show concentric circle plot with colored radii
``` {r function definition: show concentricSemiCircle-Tumor-center_coloredRadii plot, fig.width=12, fig.height=10}
# Check list of outputs saved under res
names(res)

# Show plot with concentric circles emanating from Tumor center
res$concCircle.plot.radius

```


# Function to annotate metadata with radius bins

## Function Definition
``` {r function definition: annotate Xenium metadata with radius bins, eval = FALSE}
This function annotate Xenium metada with radius groups.

annotate_radius_bins_xenium <- function(xenium.obj, data, concentric_circles_sf, save_output = TRUE, metadata_file = "meta.xenium.merge.rds", xeniumObj_file="XeniumfullObject_sfConcCircles.rds")

  
# Params:
#' @param xenium.obj A Seurat Xenium object containing cell metadata.
#' @param data Data frame containing spatial coordinates and cell metadata.
#' Must contain columns \code{cell_ID}, \code{x.coord}, and
#' \code{y.coord}.
#' @param concentric_circles_sf An sf object containing concentric circle
#' geometries generated by \code{generate_concentric_circles()}.
#' @param save_output Logical. If TRUE, save the annotated metadata and
#' updated Xenium object as RDS files.
#' @param metadata_file Character string specifying the filename used to save
#' merged metadata.
#' @param xeniumObj_file Character string specifying the filename used to save
#' the updated Xenium object.



The annotate_radius_bins_xenium() function assigns each cell to the smallest concentric circle radius that contains it by performing a spatial join between cell coordinates and the concentric circle geometries generated by generate_concentric_circles_xenium(). The resulting radius assignments are merged back into the parent Giotto object as a new metadata column (radius.concentric.sf), enabling downstream spatial analyses based on distance from the tumor centroid. This function serves as the critical bridge between the concentric circle geometries and cell-level metadata, allowing each cell to be associated with a specific radius group.


Output files:
xenium_cells_annotated: An sf object containing the original cell coordinates together with the radius assignment obtained from the spatial join.
xenium_cells_clean_binned: A cleaned metadata table containing a single radius assignment per cell after removing duplicate assignments and retaining the smallest matching radius.
meta.xenium.merge: A merged metadata table containing all cells and their corresponding radius.concentric.sf annotations. This table is used as the primary input for downstream radius-based analyses.
xenium.obj: An updated Xenium object with the new radius.concentric.sf metadata added to each cell.

When save_output = TRUE, the merged metadata is automatically saved as an RDS file (meta.xenium.merge.rds by default) and the updated Xenium object is saved as an RDS file (XeniumfullObject_sfConcCircles.rds by default). These saved objects allow users to resume downstream analyses without repeating the computationally intensive spatial join step.

The outputs generated by annotate_radius_bins_xenium() are subsequently used by functions such as label_peritumoral_distal_xenium(), gene_count_radius_xenium().

```


## Deploy function to annotate metadata with radius bins
``` {r function deployment: annotate Xenium metadata with radius bins}

res <- annotate_radius_bins_xenium(xenium.obj = xenium.obj, data = res.conc$xenium.sub, concentric_circles_sf = res.conc$concentric_circles_sf, save_output = TRUE, metadata_file = "meta.xenium.merge.rds", xeniumObj_file="XeniumfullObject_sfConcCircles.rds")


```


# Retrieve previously saved meta.xenium.merge file
``` {r function deployment: retrieve meta.xenium.merge}
# Make sure to load previously saved metadata files for downstream functions

# Output file from annotate_radius_bins_giotto()
meta.xenium.merge <- readRDS("meta.xenium.merge.rds")


# meta.xenium.merge: A merged metadata table containing all cells and their corresponding radius.concentric.sf annotations. This table is used as the primary input for downstream radius-based analyses.


```


# Function to plot radius groups as peritumoral vs distal regions

## Function Definition
``` {r function definition: label radius groups as peritumoral vs distal regions, eval = FALSE}
This function categorizes spatial radius groups into distinct microenvironments relative to the tumor core:
"Tumor" (core), "Peritumoral" (immediately adjacent), and "Distal" (distant/stromal).


label_peritumoral_distal_xenium <- function(data, meta.xenium.merge, tumor.radius=1000, peritumoral.radius=2800, distal.radius=7400, concentric_circles_sf, center_point_sfc, g.title="Concentric Circles with Colored Boundaries", tumor.conc.color="darkred", peritumoral.conc.color="goldenrod", distal.conc.color="darkolivegreen4", save_output = TRUE, save_plot = TRUE, plot_file = "concentricSemiCircle_peritumoralDistalGroupColored.png")
  
  
# Params:
#' @param data Data frame containing spatial coordinates. Typically
#'   \code{res.conc$xenium.sub}.
#' @param meta.xenium.merge Data frame containing concentric circle radius
#'   annotations for each cell.
#' @param tumor.radius Radius defining the tumor boundary.
#' @param peritumoral.radius Radius defining the peritumoral boundary.
#' @param distal.radius Radius defining the distal boundary.
#' @param concentric_circles_sf sf object containing concentric circle
#'   geometries generated by \code{generate_concentric_circles()}.
#' @param center_point_sfc sf point object representing the tumor centroid.
#' @param g.title Plot title.
#' @param tumor.conc.color Color used for the tumor boundary.
#' @param peritumoral.conc.color Color used for the peritumoral boundary.
#' @param distal.conc.color Color used for the distal boundary.
#' @param save_output Logical; whether to save generated output
#'   (RDS).
#' @param save_plot Logical; whether to save generated plot
#'   (PNG plot).
#' @param plot_file Output PNG filename used when
#'   \code{save_plot = TRUE}.


The label_peritumoral_distal_xenium() function converts concentric circle radius assignments into biologically meaningful spatial regions by defining Tumor, Peritumoral, and Distal boundaries around the tumor centroid. Using the radius annotations generated by annotate_radius_bins_xenium(), the function assigns each radius group to a broader spatial compartment, generates region-level metadata for downstream analyses, and creates a visualization highlighting the selected boundary rings. This step simplifies the interpretation of concentric circle analyses by grouping multiple radius bins into biologically relevant tumor-proximal and tumor-distal regions.


Output files:
meta.merge.peri: A metadata table containing radius assignments, radius group labels, and higher-level spatial region annotations (Tumor, Peritumoral, and Distal). This table serves as the primary metadata input for downstream spatial analyses.
label.check: A quality-control summary table used to verify that radius groups were correctly assigned to their corresponding spatial regions.
data.merge: Cell metadata merged with the newly assigned spatial region labels.
radius_map: A lookup table linking concentric circle radii to their corresponding spatial regions.
boundary_sf: An sf object containing only the Tumor, Peritumoral, and Distal boundary rings used for visualization.
concCircle.plot.radius: A ggplot object displaying the spatial boundaries overlaid on the tissue coordinates.

When save_output = TRUE, the annotated metadata is automatically saved as an RDS file (meta.merge.peri.rds by default). When save_plot = TRUE, a high-resolution PNG file (concentricSemiCircle_peritumoralDistalGroupColored.png by default) is generated showing the Tumor, Peritumoral, and Distal boundaries highlighted using user-defined colors. This visualization serves as an important quality-control step to verify that the selected boundary radii correctly capture the intended spatial regions before performing downstream analyses.

The outputs generated by label_peritumoral_distal_xenium() are commonly used in downstream workflows such as gene_count_radius_xenium(), differential expression analyses, cell-type enrichment studies, and tumor microenvironment characterization, where comparisons are made between tumor-proximal and tumor-distal spatial compartments.

```

## Deploy function to plot radius groups as peritumoral vs distal regions
``` {r function deployment: label radius groups as peritumoral vs distal regions}

res <- label_peritumoral_distal_xenium(data = res.conc$xenium.sub, meta.xenium.merge = meta.xenium.merge, tumor.radius=1000, peritumoral.radius=2800, distal.radius=7400, concentric_circles_sf, center_point_sfc, g.title="Concentric Circles with Colored Boundaries", tumor.conc.color="darkred", peritumoral.conc.color="goldenrod", distal.conc.color="darkolivegreen4", save_output = TRUE, save_plot = TRUE, plot_file = "concentricSemiCircle_peritumoralDistalGroupColored.png")

```

## Show previously plotted peritumoral vs distal regions
``` {r show peritumoral vs distal regions, fig.width=12, fig.height=10}
# Check list of outputs saved under res
names(res)

# Show plot with peritumoral vs distal regions
res$concCircle.plot.radius

```


# Retrieve previously saved meta.merge.peri file
``` {r function deployment: retrieve meta.merge.peri}
# Make sure to load previously saved metadata files for downstream functions

# Output file from label_peritumoral_distal_giotto()
meta.merge.peri <- readRDS("meta.merge.peri.rds")


# meta.merge.peri: A metadata table containing radius assignments, radius group labels, and higher-level spatial region annotations (Tumor, Peritumoral, and Distal). This table serves as the primary metadata input for downstream spatial analyses.


```


# Function to average raw gene count per radius group

## Function Definition
``` {r function function: calculate raw gene count per radius group, eval = FALSE}
This function calculates aggregated raw gene expression ("Raw" slot in Giotto object) per radius group.
It further normalizes the raw gene expression signal by the total cell count within each radius group.
It automatically generates gene count plots for the customized gene list.

gene_count_radius_xenium <- function(xenium.obj, meta.xenium, sample, select.genes, count.type=c("Raw","Normalized"), plot1.color="darkblue", plot2.color="darkgreen")


# Params:
#' @param xenium.obj A Seurat Xenium object containing expression data and
#'   metadata.
#' @param meta.xenium Data frame containing cell metadata. Must include
#'   \code{cell_ID} and \code{radius_group_label} columns.
#' @param sample Character string specifying the sample name to analyze.
#'   Must match values stored in \code{orig.ident}.
#' @param select.genes Character vector of gene names to summarize and plot.
#' @param count.type Type of expression matrix to use. One of:
#'   \itemize{
#'     \item{"Raw"} Raw transcript counts (\code{counts} layer).
#'     \item{"Normalized"} Normalized expression values (\code{data} layer).
#'   }
#' @param plot1.color Color used for plots showing total gene counts per
#'   radius group.
#' @param plot2.color Color used for plots showing gene counts normalized
#'   by total cells per radius group.


The gene_count_radius_xenium() function quantifies gene expression patterns across concentric radius groups generated from tumor-centered spatial regions. Using either raw transcript counts (count.type = "Raw") or normalized expression values (count.type = "Normalized"), the function aggregates expression values for a user-defined gene list and visualizes how gene abundance changes as a function of distance from the tumor centroid. For each selected gene, the function generates two complementary summaries: 
(1) total gene counts per radius group and 
(2) gene counts normalized by the total number of cells within each radius group. These visualizations provide a straightforward way to identify spatial expression gradients and compare gene enrichment across tumor-proximal and tumor-distal regions.


Output files:
seu: A lightweight Seurat object containing only the subsetted expression matrix and corresponding metadata. This object can be used for downstream exploratory analyses without requiring the full Xenium object.

For each gene listed in select.genes, the function automatically generates and saves two publication-ready PNG files:

Gene_count_by_radius_group_[GENE].png: Displays the total gene counts aggregated across each radius group.
Gene_count_by_radius_group[normTotalCellsPerGroup]_[GENE].png: Displays gene counts normalized by the total number of cells within each radius group, allowing direct comparison of expression intensity independent of cell abundance.

When the function is executed, a lightweight Seurat object is also automatically saved as an RDS file (Seurat_XeniumObj_RawCounts.RDS or Seurat_XeniumObj_NormalizedCounts.RDS, depending on the selected count.type). This allows users to quickly reload the subsetted expression matrix and metadata without repeatedly extracting expression data from the Xenium object.

These outputs are typically generated after running annotate_radius_bins_xenium() and label_peritumoral_distal_xenium(), which provide the radius group annotations required for aggregation. The resulting plots can be used to investigate spatial gene expression gradients, evaluate immune cell localization patterns, identify tumor-associated transcriptional programs, and prioritize candidate biomarkers for downstream analyses.

```

## Deploy function to raw average gene count per radius group
``` {r function deployment: calculate raw gene count per radius group}

gene_count_radius_xenium(xenium.obj = xenium.obj, meta.xenium = meta.merge.peri, select.genes =c("FOXP3", "IL2RA", "CD4", "LAG3", "HAVCR2", "ENTPD1"), count.type="Raw", plot1.color="darkblue", plot2.color="darkgreen")

```

## Show raw gene count plots
``` {r show raw gene count plots, echo=FALSE, fig.width=12, fig.height=10}

knitr::include_graphics(
  "Gene_count_by_radius_group_Raw_FOXP3.png"
)

knitr::include_graphics(
  "Gene_count_by_radius_group[normTotalCellsPerGroup]_Raw_FOXP3.png"
)

```

# Function to average normalized gene count per radius group

## Function Definition
``` {r function definition: calculate normalized gene count per radius group, eval = FALSE}

# Instead of raw gene count, this function also calculates aggregated normalized gene expression ("Normalized" slot in Giotto object) per radius group.
# It further normalizes the raw gene expression signal by the total cell count within each radius group.
# It automatically generates gene count plots for the customized gene list.

gene_count_radius_xenium <- function(xenium.obj, meta.xenium, select.genes, count.type="Normalized", plot1.color="darkblue", plot2.color="darkgreen")

  
# Params:
#' @param xenium.obj A Seurat Xenium object containing expression data and
#'   metadata.
#' @param meta.xenium Data frame containing cell metadata. Must include
#'   \code{cell_ID} and \code{radius_group_label} columns.
#' @param sample Character string specifying the sample name to analyze.
#'   Must match values stored in \code{orig.ident}.
#' @param select.genes Character vector of gene names to summarize and plot.
#' @param count.type Type of expression matrix to use. One of:
#'   \itemize{
#'     \item{"Raw"} Raw transcript counts (\code{counts} layer).
#'     \item{"Normalized"} Normalized expression values (\code{data} layer).
#'   }
#' @param plot1.color Color used for plots showing total gene counts per
#'   radius group.
#' @param plot2.color Color used for plots showing gene counts normalized
#'   by total cells per radius group.


The gene_count_radius_xenium() function quantifies gene expression patterns across concentric radius groups generated from tumor-centered spatial regions. Using either raw transcript counts (count.type = "Raw") or normalized expression values (count.type = "Normalized"), the function aggregates expression values for a user-defined gene list and visualizes how gene abundance changes as a function of distance from the tumor centroid. For each selected gene, the function generates two complementary summaries: 
(1) total gene counts per radius group and 
(2) gene counts normalized by the total number of cells within each radius group. These visualizations provide a straightforward way to identify spatial expression gradients and compare gene enrichment across tumor-proximal and tumor-distal regions.


Output files:
seu: A lightweight Seurat object containing only the subsetted expression matrix and corresponding metadata. This object can be used for downstream exploratory analyses without requiring the full Xenium object.

For each gene listed in select.genes, the function automatically generates and saves two publication-ready PNG files:

Gene_count_by_radius_group_[GENE].png: Displays the total gene counts aggregated across each radius group.
Gene_count_by_radius_group[normTotalCellsPerGroup]_[GENE].png: Displays gene counts normalized by the total number of cells within each radius group, allowing direct comparison of expression intensity independent of cell abundance.

When the function is executed, a lightweight Seurat object is also automatically saved as an RDS file (Seurat_XeniumObj_RawCounts.RDS or Seurat_XeniumObj_NormalizedCounts.RDS, depending on the selected count.type). This allows users to quickly reload the subsetted expression matrix and metadata without repeatedly extracting expression data from the Xenium object.

These outputs are typically generated after running annotate_radius_bins_xenium() and label_peritumoral_distal_xenium(), which provide the radius group annotations required for aggregation. The resulting plots can be used to investigate spatial gene expression gradients, evaluate immune cell localization patterns, identify tumor-associated transcriptional programs, and prioritize candidate biomarkers for downstream analyses.

```

## Deploy function average normalized gene count per radius group
``` {r function deployment: calculate normalized gene count per radius group}

gene_count_radius_xenium(xenium.obj = xenium.obj, meta.xenium = meta.merge.peri, select.genes =c("FOXP3", "IL2RA", "CD4", "LAG3", "HAVCR2", "ENTPD1"), count.type="Normalized", plot1.color="darkblue", plot2.color="darkgreen")

```


## Show normalized gene count plots
``` {r show normalized gene count plots, echo=FALSE, fig.width=12, fig.height=10}

knitr::include_graphics(
  "Gene_count_by_radius_group_Normalized_FOXP3.png"
)

knitr::include_graphics(
  "Gene_count_by_radius_group[normTotalCellsPerGroup]_Normalized_FOXP3.png"
)

```

